---
title: "Assignment 1"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Student #: 23088091  
Name : Sean Peralta Garcia


# 1. Executive Summary

# 2. Introduction
* talk about literature reviews
* explain how i think that the house prices in saratoga will largely follow the patterns seen in other places
* explain my data - prices and variables from houses in saratoga and what - all details about property
  + assumed that the prices of the properties are taken at the same time (demand pressures)
  + data is only intrinsic to property, no location data - same neighbourhood so assume this is a non factor
  + different types of independent variables
* state my hypothesis and aims of analysis

# 3. Methodology
* plot() - plots all combinations of terms
  + only use some in results - ones of note

* plot measures with different factors and if the least squares difference line crosses then it has interaction
--------
* mathematical justification for choosing transformation
* find relationship approx linear
* with approx constant variance

* scatter plot to see if there are any obvious transformations
  + and least squares regression line to find outliers (find points above and below line)
  + residuals vs fitted and select outliers more than 2std deviations
* fit a linear model wk2 and 3 lab - no r code shown
* residual plot to see what variables should be transformed
* estimated regression line
* cross table
* residual plot

# 4. Results

# 5. Discussion

# 6. References


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
